Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation

### 2025-04-11 manuscript/manuscript.qmd
updated manuscript

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigation
      - Software
      - Visualization
    affiliations:
      - Consultant
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Investigation
      - Project administration
    affiliations:
      - NASA
abstract: Placeholder
plain-language-summary: Placeholder
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: references.bib
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

@@

Some researchers have attempted to deal with these issues. [@seegers2018] e.g. have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach. Others have tried to go a step futher and switch to the Bayesian paradigm. E.g. [ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction. [@shi2015] proposed a probabilistic method to merge data from different sensors. [@craig2019] have proposed a Bayesian neural network approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance. [@werther2022] used Monte-Carlo dropout as deep neural net training method to obtain prediction uncertainty. [@erickson2023] have proposed using conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a forward model. [@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations. Yet most of these approaches retain variable levels of frequentism by applying only part of what is commonly referred to as the Bayesian workflow[@betancourt2019].

In this paper we propose a modified workflow to Bayesian modeling for Ocean Color emote sensing. We propose that this workflow include; (1) one or more model building, (2) prior formulation via expert knowledge elicitation and/or data-free simulation, (3) fitting model to data, (4) model validation, comparison and selection, (5) interpretation of results via posterior analysis, (6) prediction on new observation, (7) update the posterior when new labeled data becomes available. As illustration we recast some of the pre-existing $OCx$ models into their probabilistic version. To illustrate model comparison and selection we also build and evaluate Bayesian Additive Tree Regregssion  (BART; [@chipman2010bart]) type model.

>>>>

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independant Consultant
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA
abstract: Placeholder
plain-language-summary: Placeholder
keywords: []
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: references.bib
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

@@

#### Overcoming limitations
Some researchers have attempted to deal with these issues. [@seegers2018] e.g. have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach. Others have tried to go a step futher and switch to the Bayesian paradigm. E.g. [ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction. [@shi2015] proposed a probabilistic method to merge data from different sensors. [@craig2019] have proposed a Bayesian neural network approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance. [@werther2022] used Monte-Carlo dropout as deep neural net training method to obtain prediction uncertainty. [@erickson2023] have proposed using conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a forward model. [@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations. Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].

#### In search of the Ocean Color Bayesian workflow
The Bayesian workflow is a recognizable and unifying way of approaching statistical modeling, initially championed by statisticians at Columnbia University [@gelman_bda] subsequently quick endorsed and further developed by researchers dealing with scant and noisy data (e.g. [@mcelreath_stat_rethink]). That is not to say that all steps are applicable to all fields of research and all intents. The workflow can and should be tailored to the researcher's field. [@wolkovich2024fourstepbayesianworkflowimprovin] has proposed a four-step process for Ecological modeling. 
However there are some core steps that should be followed. (1) Begin with one, or better yet,  more than one conceptual models and code them up.  Model building include prior formulation to encode existing knowledge. Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvemnt. (2) Check assumptions by a simulation process known as *Prior Predictive Checks*. Bayesian models are generative, meaning they can run on empty, that is produce results without data. This allows a sanity check of the built-in assumptions. Non-sensical simulation output indicate assumptions must be revisited. (3) Collect data and conduct exploratory data analysis. Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features. (4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information. Most modern packages (e.g. Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights. (5) Model compasison and selection. Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights. (6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.

In this paper, we largely follow this workflow. We begin by leveraging some of the work eluded to above and  recast some of the pre-existing $OCx$ models into their probabilistic version. We also offer an alternative in the form of Bayesian Additive Regression Trees  (BART; [@chipman2010bart]).

### 2025-04-11 manuscript/manuscript.tex
updated manuscript

Satellite ocean color observations have long been fundamental for
monitoring marine ecosystems, as they enable the global estimation of
chlorophyll‑a---a key indicator of phytoplankton biomass and ocean
productivity. Early empirical algorithms, notably developed by O'Reilly
et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al., 2000),
established the \(OCx\) family (where \(x\) denotes the number of bands
used) of polynomial regression models. These models relate blue-to-green
reflectance ratios (after log‑transformation) to in situ chlorophyll‑a,
employing either straight band ratios (BR) or maximum band ratios
(MBR)---the latter selecting the highest available blue-to-green ratio
for any given observation as input to a high‑order polynomial. These
formulations have served as the operational foundation for chlorophyll‑a
products across a broad range of satellite ocean color sensors---from
the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS,
and MERIS to more recent missions---offering a straightforward and
robust approach for Case‑1 waters. However, their performance is more
limited in optically complex Case‑2 waters and remains sensitive to
atmospheric correction errors.

Subsequent refinements were introduced to address these deficiencies.
For example, Hu et al. (Hu et al., 2012) proposed a Color Index (CI)
formulation that employs a band‑difference approach to reduce
sensitivity to residual atmospheric errors and instrument noise, with
further improvements enhancing inter‑sensor agreement (Hu et al., 2019).
The increasing availability of calibration data (e.g.,
(\textbf{Valente2015?})) and ongoing algorithmic improvements have led
to the development of additional variants of the \(OCx\)
algorithms---specifically, the OC5 and OC6 formulations. O'Reilly and
Werdell (John E. O'Reilly \& Werdell, 2019) maintain that OC5 extends
the spectral basis by incorporating the 412\,nm band, thereby exploiting
its strong signal in clear, oligotrophic waters, while OC6 replaces the
traditional denominator with the mean of the 555 and 670\,nm
reflectances, improving the dynamic range at low chlorophyll
concentrations. In total, (John E. O'Reilly \& Werdell, 2019) propose 65
versions of BR/MBR \(OCx\) type algorithms for 25 sensors---on average,
two or more variants per sensor. With this arsenal, it is hoped,
researchers are better equipped to address the wide array of bio‑optical
environments encountered in global ocean color applications.

@@

Regrettably, the development of traditional ocean color algorithms
relies on a fundamental statistical error. This error originates in the
conflation of empirical frequencies --- derived from a limited, noisy
set of observations --- with the theoretical relative frequencies that
would be observed over an infinite number of trials, thereby ignoring
the intrinsic uncertainty and variability inherent in finite sampling.
The fallout has and continues to be extremely detrimental to
observational science. Traditional, frequentist approaches, where the
variability in sample frequencies is treated as negligible, yield
overconfident inferences, and allegedgly significant findings that do
not hold water, in time leading to replication crises.

In statistical modeling, the goal is to assess how well a hypothesized
model \(M\) represents the relationship between independent and
dependent variables based on collected data \(D\). Frequentist methods
focus exclusively on the likelihood \(p(D|M)\), estimating model
parameters by maximizing this probability. This approach assumes that,
with a sufficiently large dataset, the chosen parameters reliably
reflect the true system. However, in practice, any finite dataset
carries inherent uncertainty and noise. As a consequence, models
developed under this framework tend to be overconfident - yielding point
estimates that do not adequately capture the variability of
\emph{in-situ} observations - and may not generalize well to new data.
Ad hoc attempts to generate uncertainty, such as bootstrapping, further
compound the problem by resampling the flawed, finite observations and
thus amplifying any existing biases. This overconfidence, coupled with
unverified assumptions presented as objectivity, can lead to misleading
inferences and contribute to replication challenges. The historical
process by which these assumptions have been enshrined is fascinating in
its own right, and we invite interested readers to consult works such as
McGrayne (\textbf{mcgrayne?}) and Clayton (\textbf{clayton?}).

As elaborated extensively by Jaynes (\textbf{jaynes2003probability?}),
and more recently by practitioners like Gelman
(\textbf{gelman2013bayesian?}), and McElreath
(\textbf{mcelreath2015statistical?}), probability is best understood as
a degree of knowledge rather than as a long-run frequency. Accordingly
in this Bayesian framework, the prior \(p(M)\) encodes our current state
of knowledge about the model's parameters before \(D\) is observed. This
knowledge can be elicited from experts, or if none exixts weakly
informative priors can be used. Note that whether \(M\) is based on
empirically observed relationships, analytical understanding, or
something in between is irrelevant to its integration into this
framework. Once the data \(D\) are collected, Bayes' theorem combines
the likelihood \(p(D|M)\) with the prior \(p(M)\) to yield the posterior
distribution \(p(M|D)\), which represents an updated level of knowledge
about the model's parameters. Crucially this posterior can then serve as
the new prior when new data become available, allowing for sequential
integration of new information. The by-product of this approach is
uncertainty estimates that more accurately reflect the complexities of
the underlying system as currently understood. In effect, rather than
relying on single point estimates based on \(p(D|M)\), the posterior
\(p(M|D)\) can be mined to retrieve a range of plausible parameter
values through credible intervals offering a more transparent and robust
basis for inference. This principled integration of prior knowledge with
observed data is central to our approach, as detailed further in the
Supplementary Material.

In this paper we recast a number of OCx-like algorithms in the Bayesian
framework. We also explore a different kind of

\subsubsection{Motivation for Bayesian Tree-Based
Models}\label{motivation-for-bayesian-tree-based-models}

To address these limitations, we explore Bayesian tree-based modeling --
specifically, Bayesian Additive Regression Trees (BART) -- as a novel
approach for ocean color chlorophyll retrieval. BART offers several
advantages for this application:

Direct Bayesian inference: As a Bayesian ensemble method, BART directly
targets the posterior distribution \(p(Chl | Rrs)\) rather than a purely
empirical fit. The model inherently provides a probabilistic mapping
from reflectance to chlorophyll. Rather than producing a single best-fit
value, BART returns a posterior mean and credible interval for each
prediction, naturally quantifying predictive uncertainty.

Uncertainty quantification: The BART framework includes uncertainty in
its predictions by design. Every estimated chlorophyll value comes with
a confidence range (e.g., 95\% credible interval) reflecting model
uncertainty and observational noise. This is a key improvement over
deterministic algorithms, enabling more reliable use of the data for
scientific and operational purposes where understanding uncertainty is
crucial e.g.~for forecasting trends.

Interpretability: Being a sum-of-trees model, BART can be interrogated
to understand the influence of each input variable. Tools such as
partial dependence plots allow us to visualize the relationship between
each spectral band and the predicted chlorophyll, marginalizing over
other bands. In contrast to high-order polynomial coefficients, these
tree-based partial dependencies are intuitively interpretable. For
instance, one can observe how changes in a particular waveband's
reflectance (holding others average) drive the chlorophyll estimate,
revealing any thresholds or non-linear responses.

BART's modeling approach is non-parametric and highly flexible, which
means it can capture complex, non-linear relationships between
multi-spectral Rrs inputs and chlorophyll without imposing a rigid
functional form. At the same time, its Bayesian regularization -- which
shrinks the contribution of each individual tree -- helps prevent
overfitting even with a rich model structure. This balance between
flexibility and regularization is well-suited to global ocean color
data, where the true Rrs--Chl relationship is known to be non-linear and
context-dependent, yet a model must generalize across diverse water
types.

The interpretability of the BART model is illustrated by partial
dependence analysis on the six input Rrs bands (centered approximately
at 412, 443, 490, 510, 555, and 670 nm). The partial dependence plots
reveal nuanced spectral--chlorophyll relationships and clear transitions
in the dominant wavelength influence across different chlorophyll
regimes. For example, the shorter blue wavelengths (412--443 nm) show a
strong inverse influence on predicted \(log(Chl_a)\) (higher blue
reflectance corresponds to lower chlorophyll) that gradually saturates
in the most oligotrophic waters, consistent with the leveling off of
blue-to-green ratios at very low Chl. The green band (\textasciitilde555
nm) exhibits a more complex, non-monotonic effect: at low reflectance
(clear water) its influence on the Chl estimate is minimal, but it
becomes increasingly important through intermediate reflectance ranges
-- reflecting the band-difference signal exploited by the CI algorithm
-- and then diminishes for very high chlorophyll where green reflectance
tends to flatten. In contrast, the red band (\textasciitilde670 nm) has
virtually no effect on the model's predictions until a threshold is
reached at elevated chlorophyll concentrations, after which its
influence steeply increases. This behavior aligns with optical
expectations, since red reflectance is negligible in low-Chl waters but
rises sharply once phytoplankton absorption in the blue saturates and
biomass is high. These partial dependence results indicate that the BART
model automatically learns the piecewise spectral logic that
oceanographers often handle via separate algorithms (blue/green ratios
for low-to-moderate Chl, red bands for high Chl). The ability to capture
such regime-dependent responses in a single unified model, and to
visualize them, underscores the interpretability and scientific insight
provided by the BART approach.

@@

@@

>>>>

Satellite ocean color observations have long been fundamental for
monitoring marine ecosystems, as they enable the global estimation of
chlorophyll‑a (\(Chl_a\)) ---a key indicator of phytoplankton biomass
and ocean productivity. Early empirical algorithms, notably developed by
O'Reilly et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al.,
2000), established the \(OCx\) family (where \(x\) denotes the number of
bands used) of polynomial regression models. These models relate
blue-to-green reflectance ratios (after log‑transformation) to in situ
\(Chl_a\), employing either straight band ratios (BR) or maximum band
ratios (MBR)---the latter selecting the highest available blue-to-green
ratio for any given observation as input to a high‑order polynomial.
These formulations have served as the operational foundation for
chlorophyll‑a products across a broad range of satellite ocean color
sensors---from the pioneering Coastal Zone Color Scanner (CZCS) through
SeaWiFS, MODIS, and MERIS to more recent missions---offering a
straightforward and robust approach for Case‑1 waters. However, their
performance is more limited in optically complex Case‑2 waters and
remains sensitive to atmospheric correction errors.

Subsequent refinements were introduced to address these deficiencies.
For example, Hu et al. (Hu et al., 2012) proposed a Color Index (CI)
formulation that employs a band‑difference approach to reduce
sensitivity to residual atmospheric errors and instrument noise, with
further improvements enhancing inter‑sensor agreement (Hu et al., 2019).
The increasing availability of calibration data (e.g.,
(\textbf{Valente2015?})) and ongoing algorithmic improvements have led
to the development of additional variants of the \(OCx\)
algorithms---specifically, the OC5 and OC6 formulations. O'Reilly and
Werdell (John E. O'Reilly \& Werdell, 2019) maintain that OC5 extends
the spectral basis by incorporating the 412\,nm band, thereby exploiting
its strong signal in clear, oligotrophic waters, while OC6 replaces the
traditional denominator with the mean of the 555 and 670\,nm
reflectances, with the aim of improving the dynamic range at low
chlorophyll concentrations. In total, (John E. O'Reilly \& Werdell,
2019) propose 65 versions of BR/MBR \(OCx\) type algorithms for 25
sensors---on average, two or more variants per sensor. With this
arsenal, it is hoped, researchers are better equipped to address the
wide array of bio‑optical environments encountered in global ocean color
applications.

@@

Regrettably, the development of traditional ocean color predictive
models relies on a fundamental statistical error that plagues most of
observational science today; that of conflating sampling probability,
with inferential probability. Consider data \(D\) that include
predictive features (e.g.~Remote sensing reflectance - Rrs) and
prediction targets (e.g.~\(Chl_a\) or phytoplankton absorption), and a
model \(M\) (e.g.~OCx) hypothesized to adequately represent the
statistical association between the two. The sampling probability
\(p(D|M)\) also referred to as the likelihood, is the probability of
\(D\) conditioned on \(M\) being ``true''. The common model fitting
practice is to maximize \(p(D|M)\) - likelihood maximization - achieved
by tuning \(M\)'s parameters. The unspoken assumption is that making
\(M\) fit \(D\) as well as possible yields the best version of \(M\)
given \(D\) i.e.~maximizing \(p(M|D)\), the inferential probability.
This sneaky swapping from \(p(D|M)\) to \(p(M|D)\) is a violation of the
rules of conditional probability, Bayes' theorem in particular.
Jaynes(\textbf{jaynes2003?}) and later Clayton through his highly
readable account (\textbf{clayton?}) are some of the more notable
whistleblowers. Occasionally, in simple or well-behaved problems, with a
large amount of data and no prior information available the maxima of
both \(p(M|D)\) and \(p(D|M)\) can indeed converge on the same answer.
Yet there is no guarantee predictability for this outcome however, and
the fallout of this abuse is non-negligible, most noticeably in terms of
models that don't generalize well, ad-hoc uncertainty estimation
attempts, and more broadly a replication crisis.

Some researchers have attempted to deal with these issues. (Seegers et
al., 2018) e.g.~have proposed alternative metrics to circumvent the
inadequate assumptions of the frequentist approach. Others have tried to
go a step futher and switch to the Bayesian paradigm. E.g. (Frouin \&
Pelletier, 2013) have proposed a Bayesian inversion scheme for
atmospheric correction. (\textbf{shi2015?}) proposed a probabilistic
method to merge data from different sensors. (Craig \& Karaköylü, 2019)
have proposed a Bayesian neural network approach using Hamiltonian Monte
Carlo sampling to retrieve Inherent Optical Properties (IOP) from
Top-of-the-atmosphere (TOA) radiance. (\textbf{werther2022?}) used
Monte-Carlo dropout as deep neural net training method to obtain
prediction uncertainty. (Erickson et al., 2023) have proposed using
conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a
forward model. (\textbf{hammout2024?}) have proposed a Bayesian Neural
Network trained using Stochastic Variational Inference to predict
\(Chl_a\) from ocean color observations. Yet most of these approaches
retain variable levels of frequentism by applying only part of what is
commonly referred to as the Bayesian workflow(\textbf{betancourt2019?}).

In this paper we propose a modified workflow to Bayesian modeling for
Ocean Color emote sensing. We propose that this workflow include; (1)
one or more model building, (2) prior formulation via expert knowledge
elicitation and/or data-free simulation, (3) fitting model to data, (4)
model validation, comparison and selection, (5) interpretation of
results via posterior analysis, (6) prediction on new observation, (7)
update the posterior when new labeled data becomes available. As
illustration we recast some of the pre-existing \(OCx\) models into
their probabilistic version. To illustrate model comparison and
selection we also build and evaluate Bayesian Additive Tree Regregssion
(BART; (Chipman et al., 2010)) type model.

\subsection{Methods}\label{methods}

\subsubsection{Prior Elicitation and Model
Formulation}\label{prior-elicitation-and-model-formulation}

\subsubsection{Data Preprocessing}\label{data-preprocessing}

\subsubsection{}\label{section}

@@

\bibitem[\citeproctext]{ref-chipman2010bart}
Chipman, H. A., George, E. I., \& McCulloch, R. E. (2010). BART:
Bayesian additive regression trees. \emph{The Annals of Applied
Statistics}, \emph{4}(1), 266--298.
\url{https://doi.org/10.1214/09-AOAS285}

\bibitem[\citeproctext]{ref-craig2019}
Craig, S. E., \& Karaköylü, E. M. (2019). Bayesian models for deriving
biogeochemical information from satellite ocean color.
\emph{EarthArXiv}. \url{https://doi.org/10.31223/osf.io/shp6y}

\bibitem[\citeproctext]{ref-erickson2023}
Erickson, Z. K., McKinna, L. I. W., Werdell, P. J., \& Cetinić, I.
(2023). Bayesian approach to a generalized inherent optical property
model. \emph{Optics Express}, \emph{31}, 22790--22801.
\url{https://doi.org/10.1364/oe.486581}

\bibitem[\citeproctext]{ref-frouin2013}
Frouin, R., \& Pelletier, B. (2013). Bayesian methodology for ocean
color remote sensing. Retrieved from
\url{https://hal.archives-ouvertes.fr/hal-00822032}

@@

\bibitem[\citeproctext]{ref-seegers2018}
Seegers, B. N., Stumpf, R. P., Schaeffer, B. A., Loftin, K. A., \&
Werdell, P. J. (2018). Performance metrics for the assessment of
satellite data products: An ocean color case study. \emph{Opt. Express},
\emph{26}(6), 7404--7422. \url{https://doi.org/10.1364/OE.26.007404}
